A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2011-11-17 papers/contact/paper.tex
compute tensor white bear densities in contact paper Tested everything.
clean up contact paper a tad Tested everything.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
\needsworklater{ We develop a functional based on FMT~\cite{roth2002whitebear}
 to ... for SAFT}
\end{abstract}

@@

\begin{figure}
\includegraphics[width=5cm]{figs/gHS-vs-n}
\caption{A comparison of the various approximations to the contact
  density.}
\label{fig:gHS-vs-n}
\end{figure}

\begin{figure}
\includegraphics[width=5cm]{figs/free-energy}
\caption{A comparison of the various approximations to the free
  energy.  Note that the White Bear functional uses a modified form of
  the Carnahan equation of state.}
\label{fig:free-energy}
\end{figure}

The hard-sphere contact density---which we will define as the density
of hard spheres touching a given hard sphere---is important for the
association term in Statistical Associating Fluid Theory (SAFT).  For
a homogeneous hard-sphere fluid, the contact density---which is to
say, $n g_{HS}(\sigma)$---is easily computed from the
Carnahan-Starling equation of state
\begin{align}
  g_{HS}(\sigma) &= \frac{1}{1-\eta} + \cdots
\end{align}
The contact density may be easily computed from a knowledge of the
contact theorem, which states that the pressure on any hard surface is
given by
\begin{align}
  p &= k_BT n_\textit{contact}
\end{align}
Since we are interested in the contact density at the hard-sphere
surface, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the free energy on hard sphere
radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
Since $\eta \equiv \frac{4\pi}{3} R^3$, we can work out that
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi (2R)^2$.  The area which is relevant
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
  &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  &= n k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
As it turns out, this is the standard answer for the contact value of
the autocorrelation function! $\ddot\smile$

@@

\section{Mean Contact density}

@@

\section{Contact density at a given hard sphere}

@@

@@

\begin{figure}
  \includegraphics[height=5cm]{figs/sphere-radial-06-04}
  \caption{Density and contact densities of four hard spheres in a
    spherical cavity with radius 6.}
  \label{fig:sphere-density}
\end{figure}

\begin{figure}
  \includegraphics[height=5cm]{figs/sphere-radial-08-13}
  \caption{Density and contact densities of fourteen hard spheres in a
    spherical cavity with radius 8.  Also similar densities of spheres
    in even larger cavities.}
  \label{fig:sphere-density2}
\end{figure}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
We get a nice spatial dependence for SAFT.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We investigate the use of the White Bear functional,
  based on Fundamental Measure Theory~\cite{roth2002whitebear} to
  predict contact densities for the inhomogeneous hard-sphere fluid.
\end{abstract}

@@

\begin{figure}
\includegraphics[width=5cm]{figs/gHS-vs-n}
\caption{A comparison of the various approximations to the contact
  density in the homogeneous limit.  As expected, they are all
  identical, and we won't include this plot in the paper (but it's
  here to verify that our code is working).}
\label{fig:gHS-vs-n}
\end{figure}

%% \begin{figure}
%% \includegraphics[width=5cm]{figs/free-energy}
%% \caption{A comparison of the various approximations to the free energy
%%   in the homogeneous limit.  Note that the White Bear functional uses
%%   a modified form of the Carnahan equation of state.  As expected,
%%   these are all identical, and we won't include this plot in the paper
%%   (but it's here to verify that our code is working).}
%% \label{fig:free-energy}
%% \end{figure}

The hard-sphere contact density---that is to say, the density of hard
spheres touching a given hard sphere---is important for the
association term in Statistical Associating Fluid Theory (SAFT).  For
a homogeneous hard-sphere fluid, the contact density---which is to
say, $n g_{HS}(\sigma)$, where $g_{HS}(\sigma)$ is the contact value
of the correlation function---is easily computed from the
Carnahan-Starling equation of state
\begin{align}
  g_{HS}(\sigma) &= \frac{1}{1-\eta} + \cdots
\end{align}
The contact density may be easily computed from a knowledge of the
contact theorem, which states that the pressure on any hard surface is
given by
\begin{align}
  p &= k_BT n_\textit{contact}
\end{align}
Since we are interested in the contact density at the hard-sphere
surface, all we need compute is the pressure on that surface, and
we'll have our answer.  The pressure on a hard sphere can be readily
computed from the dependence of the free energy on hard sphere
radius.
\begin{align}
  A_{HS} &= Nk_BT \frac{4\eta - 3\eta^2}{(1-\eta)^2}
\end{align}
Since $\eta \equiv \frac{4\pi}{3} R^3$, we can work out that
\begin{align}
  \frac{dA_{HS}}{dR} &= \frac{dA_{HS}}{d\eta} \frac{d\eta}{dR} \\
  &= Nk_BT \left( \frac{4 - 6\eta}{(1-\eta)^2} + 2 \frac{4\eta - 3\eta^2}{(1-\eta)^3} \right) \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 4\eta - 6\eta + 6\eta^2 + 8\eta - 6\eta^2}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{d\eta}{dR}
  \\
  &= Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \label{eq:dAhsdR}
\end{align}
This derivative gives us the force on \emph{all} the hard
spheres---since we're changing all their radii at once.  To compute
the pressure on the spheres, we just need to divide by the total area,
which means dividing by $N 4\pi (2R)^2$.  The area which is relevant
is the area over which the molecules can make contact.
\begin{align}
  p_{HS} &= \frac{1}{N 4\pi (2R)^2} \frac{dA_{HS}}{dR} \\
  &= \frac14 \frac{1}{N 4\pi R^2} Nk_BT \frac{4 - 2\eta}{(1-\eta)^3} \frac{3 \eta}{R} \\
  &= \frac{3}{4\pi R^3} \eta k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  &= n k_BT \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
\begin{align}
  n_\textit{contact} &= n \frac{1 - \frac{\eta}2}{(1-\eta)^3} \\
  g_{HS}(\sigma) &= \frac{1 - \frac{\eta}2}{(1-\eta)^3}
\end{align}
As it turns out, this is the standard answer for the contact value of
the autocorrelation function! $\ddot\smile$

@@

\section{Mean Contact density}\label{simple-contact}

@@

\section{Contact density at a given hard
  sphere}\label{contact-at-sphere}

@@

We performed a Monte-Carlo simulation of the hard sphere fluid to
measure the contact density as a function of position for a few simple
geometries.  For each scenario, we compute and compare several
quantities.  We compare the Monte Carlo density with the DFT
prediction.  We also plot Monte Carlo results for the two contact
densities: the contact density evaluated at contact, and the contact
density evaluated at the sphere location.  Alongside these, we plot
the contact density of Fu and Wu~\cite{fu2005vapor-liquid-dft} (or
should it be Yu and
Wu\cite{yu2002fmt-dft-inhomogeneous-associating}?), the ``simple
contact density'' from Section~\ref{simple-contact}, and the ``contact
density at sphere'' from Section~\ref{contact-at-sphere}.  In
principle, the functional of Wu (FIXME) should match the
CenConDensity, while the DFT at sphere should match the ConDensity...

@@

The first geometry we consider is a number of hard spheres confined
within a spherical cavity.  In
Figures~\ref{fig:sphere-8}-\ref{fig:sphere-16}, we show
the densities and contact densities for several cavity sizes.  In each
case, the number of hard spheres was chosen to set the mean filling
fraction to around 0.3?.  This value was chosen because it roughly
corresponds to the filling fraction of water at standard atmospheric
temperature and pressure in the optimal SAFT model developed by Clark
\emph{et al}\cite{clark2006developing}.

\textcolor{red}{We should also try looking at a few other densities.}

\newcommand\sphereExplanation{ Blue curves describe the density of
  hard spheres.  Green curves describe the contact density averaged
  according to the point of contact.  Red curves describe the contact
  density as averaged according to the centers of the spheres that are
  touching.  In each case, solid lines correspond to the Monte Carlo
  simulation, and dashed lines of various sorts represent DFT model
  predictions.  }

\begin{figure}
  \includegraphics[height=5cm]{figs/contact-08-013}
  \caption{Density and contact densities of thirteen hard spheres in a
    spherical cavity with diameter 8. \sphereExplanation }
  \label{fig:sphere-8}
\end{figure}

\begin{figure}
  \includegraphics[height=5cm]{figs/contact-12-112}
  \caption{Density and contact densities of 112 hard spheres in a
    spherical cavity with diameter 12.  \sphereExplanation}
  \label{fig:sphere-12}
\end{figure}

\begin{figure}
  \includegraphics[height=5cm]{figs/contact-16-265}
  \caption{Density and contact densities of 265 hard spheres in a
    spherical cavity with diameter 16. \sphereExplanation}
  \label{fig:sphere-16}
\end{figure}

@@

\textcolor{red}{
  I am thinking of also performing simulations for a cubical cavity, and
  examining the result near the center of a face, which (for a large
  cavity) will give us a prediction for the hard-sphere fluid near a
  single hard wall.  We've got some simulations in
  Figure~\ref{fig:box-density}, but these aren't very nice plots.
  We'll want to do a much larger box, and plot the density vs $x$ near
  the wall.  But I'm waiting on simulation data for this...
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
We (will) have compared several functionals for the contact density of
a hard-sphere fluid with simulations.  This quantity is of particular
interest, as it plays a critical role in Statistical Associating Fluid
Theory (SAFT), which is the basis of a number of recent classical
density functionals.  We compute the contact density directly from the
White Bear version of the FMT hard-sphere functional, and find
reasonable agreement?
